---
title: "Reproducible Research: Peer Assessment 1"
author: jnabonne
output: 
  html_document: 
    keep_md: yes
---

```{r knitr_global_settings, echo=F}
library(knitr)
opts_chunk$set(cache=F, warning=FALSE, message=FALSE, echo=TRUE)
```


## Loading and preprocessing the data
Checking structure after unzipping and reading the file (with correct classes)
```{r get_data}
unzip("activity.zip")
mdata <- read.csv("activity.csv", colClasses=c("numeric","Date","numeric"))
str(mdata) ; summary(mdata)
```


## What is mean total number of steps taken per day?
1. Computing the sum of steps per day (in a dataframe using dplyr)  
```{r steps_per_day}
library(dplyr)
# grouping by date then computing the total
steps_day <- mdata %>% group_by(date) %>% summarize(sumSteps=sum(steps))
```

2. Computing total number of steps **mean** and **median**
```{r mean_median_per_day}
mean(steps_day$sumSteps, na.rm=TRUE)
median(steps_day$sumSteps, na.rm=TRUE)
```

3. Printing histogram of total number of steps taken each day
```{r hist_steps_per_day}
hist(steps_day$sumSteps,
     main="histogram of the total number of steps taken each day",
     xlab="sum of steps per day")
# optionnaly printing the average of steps per day
abline(v=mean(steps_day$sumSteps, na.rm=T), lty=3, col="blue")
legend("topright", legend="mean", col="blue", lty=3)
```


## What is the average daily activity pattern?
Using dplyr again but grouping by intervals, we build the required plot  
```{r time_series_plot}
# grouping by interval then computing the average
steps_interval_avg <- mdata %>%
    group_by(interval) %>%
    summarize(meanSteps=mean(steps, na.rm=T))
plot(steps_interval_avg$meanSteps, type='l',
     main="daily average steps per 5min interval",
     xlab="interval index",
     ylab="average steps")
# optionnaly printing the max and average values as lines
abline(h=mean(steps_interval_avg$meanSteps, na.rm=T), lty=3, col="blue")
abline(h=max(steps_interval_avg$meanSteps), lty=1, col="red")
legend(x=210, y=180, legend=c("mean","max"), col=c("blue","red"), lty=c(3,1))
```

Following is the **index** and **value** of the 5-minute interval containing the maximum
number of steps (on average across all the days in the dataset):  
```{r max_interval_average}
which.max(steps_interval_avg$meanSteps)
max(steps_interval_avg$meanSteps)
```


## Imputing missing values
Checking how many observations are incomplete (_'FALSE'_ in the table below)
```{r missing_values}
table(complete.cases(mdata))
```

Creating new dataset `mdata2` with missing values imputed with interval average
across all day (previsouly computed)
```{r imputing}
mdata2 <- mdata
# parsing dataset looking for missing values...
for (i in 1:nrow(mdata2))
    if (is.na(mdata2$steps[i])) {
        # retrieving interval index in steps_interval_steps_mean
        mean_interval_idx <- steps_interval_avg$interval == mdata2$interval[i]
        # imputting na by corresponding steps_interval_avg
        mdata2$steps[i] <- steps_interval_avg$meanSteps[mean_interval_idx]
    }
```

Checking that imputed values **did not** impact dataset mean and median values
```{r checking_coherence}
# checking there is no more NA and that mean and median values did not change
summary(mdata2)
```

Checking that even if totals changed a bit, the plot global shape **has not**
```{r checking_coherence_plot}
# recomputing first step with new data set
steps_day2 <- mdata2 %>% group_by(date) %>% summarize(sumSteps=sum(steps))
# plotting the two plots side by side for easier comparison
par(mfrow=c(1,2))
hist(steps_day$sumSteps,
     main="original histo",
     xlab="steps per day")
# optionnaly printing (again) the mean (should be equal)
abline(v=mean(steps_day$sumSteps, na.rm=T), lty=3, col="blue")
legend("topright", legend="mean", col="blue", lty=3)
# the original plotting
hist(steps_day2$sumSteps,
     main="updated histogram (NA imputted)",
     xlab="steps per day")
abline(v=mean(steps_day2$sumSteps, na.rm=T), lty=3, col="blue")
legend("topright", legend="mean", col="blue", lty=3)
```


## Are there differences in activity patterns between weekdays and weekends?
First, adding a new factor variable for weekend / weekday classification
```{r days_classification}
weekend=c("Saturday", "Sunday")
mdata2 <- mdata2 %>%
    mutate(weekday=ifelse(weekdays(date) %in% weekend, "weekend", "weekday"))
```

Using Lattice to build plot displaying diff between week and weekend activity
```{r days_activity}
# grouping data per  interval AND day type (weekday) this time
steps_interval_avg2 <- mdata2 %>%
    group_by(weekday, interval) %>% summarize(meanSteps=mean(steps))
library(lattice)
xyplot(meanSteps ~ interval | weekday, data=steps_interval_avg2, type='l',
       layout=c(1,2), ylab="Number of steps",
       main="average steps per 5min interval accross week/weekend days")
```


